---
title: "GAM lake water quality trend evaluation - King County Large Lakes"
output:
  html_document:
    css: kable.css
    includes:
      in_header: cssloaders_in_header.html
self_contained: yes
runtime: shiny
resource_files:
- data/datprc.RData
- data/locs.RData
- data/map.RData
- data/mods_chl0512.RData
- data/mods_fc0512.RData
- data/mods_secchi0512.RData
- data/mods_alk0612_epi.RData
- data/mods_alk0612_hyp.RData
- data/mods_alk0612_whole.RData
- data/mods_alk0852_epi.RData
- data/mods_alk0852_hyp.RData
- data/mods_alk0852_whole.RData
- data/mods_alkA522_epi.RData
- data/mods_alkA522_hyp.RData
- data/mods_alkA522_whole.RData
- data/mods_chl0540.RData
- data/mods_chl0611.RData
- data/mods_chl0612.RData
- data/mods_chl0804.RData
- data/mods_chl0826.RData
- data/mods_chl0831.RData
- data/mods_chl0852.RData
- data/mods_chlA522.RData
- data/mods_cond0512_surface.RData
- data/mods_cond0540_surface.RData
- data/mods_cond0611_epi.RData
- data/mods_cond0611_hyp.RData
- data/mods_cond0612_epi.RData
- data/mods_cond0612_hyp.RData
- data/mods_cond0612_whole.RData
- data/mods_cond0804_surface.RData
- data/mods_cond0826_epi.RData
- data/mods_cond0826_hyp.RData
- data/mods_cond0831_epi.RData
- data/mods_cond0831_hyp.RData
- data/mods_cond0852_epi.RData
- data/mods_cond0852_hyp.RData
- data/mods_cond0852_whole.RData
- data/mods_condA522_epi.RData
- data/mods_condA522_hyp.RData
- data/mods_condA522_whole.RData
- data/mods_do0512_surface.RData
- data/mods_do0540_surface.RData
- data/mods_do0611_epi.RData
- data/mods_do0611_hyp.RData
- data/mods_do0612_epi.RData
- data/mods_do0612_hyp.RData
- data/mods_do0612_whole.RData
- data/mods_do0804_surface.RData
- data/mods_do0826_epi.RData
- data/mods_do0826_hyp.RData
- data/mods_do0831_epi.RData
- data/mods_do0831_hyp.RData
- data/mods_do0852_epi.RData
- data/mods_do0852_hyp.RData
- data/mods_do0852_whole.RData
- data/mods_doA522_epi.RData
- data/mods_doA522_hyp.RData
- data/mods_doA522_whole.RData
- data/mods_fc0540.RData
- data/mods_fcA522.RData
- data/mods_no3A522_epi.RData
- data/mods_no3A522_hyp.RData
- data/mods_no3A522_whole.RData
- data/mods_no30512_surface.RData
- data/mods_no30540_surface.RData
- data/mods_no30611_epi.RData
- data/mods_no30611_hyp.RData
- data/mods_no30612_epi.RData
- data/mods_no30612_hyp.RData
- data/mods_no30612_whole.RData
- data/mods_no30804_surface.RData
- data/mods_no30826_epi.RData
- data/mods_no30826_hyp.RData
- data/mods_no30831_epi.RData
- data/mods_no30831_hyp.RData
- data/mods_no30852_epi.RData
- data/mods_no30852_hyp.RData
- data/mods_no30852_whole.RData
- data/mods_ph0512_surface.RData
- data/mods_ph0540_surface.RData
- data/mods_ph0611_epi.RData
- data/mods_ph0611_hyp.RData
- data/mods_ph0612_epi.RData
- data/mods_ph0612_hyp.RData
- data/mods_ph0612_whole.RData
- data/mods_ph0804_surface.RData
- data/mods_ph0826_epi.RData
- data/mods_ph0826_hyp.RData
- data/mods_ph0831_epi.RData
- data/mods_ph0831_hyp.RData
- data/mods_ph0852_epi.RData
- data/mods_ph0852_hyp.RData
- data/mods_ph0852_whole.RData
- data/mods_secchi0540.RData
- data/mods_secchi0611.RData
- data/mods_secchi0612.RData
- data/mods_secchi0804.RData
- data/mods_secchi0826.RData
- data/mods_secchi0831.RData
- data/mods_secchi0852.RData
- data/mods_secchiA522.RData
- data/mods_srp0512_surface.RData
- data/mods_srp0540_surface.RData
- data/mods_srp0611_epi.RData
- data/mods_srp0611_hyp.RData
- data/mods_srp0612_epi.RData
- data/mods_srp0612_hyp.RData
- data/mods_srp0612_whole.RData
- data/mods_srp0804_surface.RData
- data/mods_srp0826_epi.RData
- data/mods_srp0826_hyp.RData
- data/mods_srp0831_epi.RData
- data/mods_srp0831_hyp.RData
- data/mods_srp0852_epi.RData
- data/mods_srp0852_hyp.RData
- data/mods_srp0852_whole.RData
- data/mods_srpA522_epi.RData
- data/mods_srpA522_hyp.RData
- data/mods_srpA522_whole.RData
- data/mods_temp0512_surface.RData
- data/mods_temp0540_surface.RData
- data/mods_temp0611_epi.RData
- data/mods_temp0611_hyp.RData
- data/mods_temp0612_epi.RData
- data/mods_temp0612_hyp.RData
- data/mods_temp0612_whole.RData
- data/mods_temp0804_surface.RData
- data/mods_temp0826_epi.RData
- data/mods_temp0826_hyp.RData
- data/mods_temp0831_epi.RData
- data/mods_temp0831_hyp.RData
- data/mods_temp0852_epi.RData
- data/mods_temp0852_hyp.RData
- data/mods_temp0852_whole.RData
- data/mods_tempA522_epi.RData
- data/mods_tempA522_hyp.RData
- data/mods_tempA522_whole.RData
- data/mods_tn0804_surface.RData
- data/mods_tn0826_epi.RData
- data/mods_tn0826_hyp.RData
- data/mods_tn0512_surface.RData
- data/mods_tn0540_surface.RData
- data/mods_tn0611_epi.RData
- data/mods_tn0611_hyp.RData
- data/mods_tn0612_epi.RData
- data/mods_tn0612_hyp.RData
- data/mods_tn0612_whole.RData
- data/mods_tn0831_epi.RData
- data/mods_tn0831_hyp.RData
- data/mods_tn0852_epi.RData
- data/mods_tn0852_hyp.RData
- data/mods_tn0852_whole.RData
- data/mods_tpA522_epi.RData
- data/mods_tnA522_hyp.RData
- data/mods_tnA522_whole.RData
- data/mods_tnA522_epi.RData
- data/mods_tp0512_surface.RData
- data/mods_tp0540_surface.RData
- data/mods_tp0611_epi.RData
- data/mods_tp0611_hyp.RData
- data/mods_tp0612_epi.RData
- data/mods_tp0612_hyp.RData
- data/mods_tp0612_whole.RData
- data/mods_tp0804_surface.RData
- data/mods_tp0826_epi.RData
- data/mods_tp0826_hyp.RData
- data/mods_tp0831_epi.RData
- data/mods_tp0831_hyp.RData
- data/mods_tp0852_epi.RData
- data/mods_tp0852_hyp.RData
- data/mods_tp0852_whole.RData
- data/mods_tpA522_hyp.RData
- data/mods_tpA522_whole.RData
- data/mods_dosat0512_surface.RData
- data/mods_dosat0540_surface.RData
- data/mods_dosat0611_epi.RData
- data/mods_dosat0611_hyp.RData
- data/mods_dosat0612_epi.RData
- data/mods_dosat0612_hyp.RData
- data/mods_dosat0612_whole.RData
- data/mods_dosat0804_surface.RData
- data/mods_dosat0826_epi.RData
- data/mods_dosat0826_hyp.RData
- data/mods_dosat0831_epi.RData
- data/mods_do0831_hyp.RData
- data/mods_dosat0831_hyp.RData
- data/mods_dosat0852_epi.RData
- data/mods_dosat0852_hyp.RData
- data/mods_do0852_whole.RData
- data/mods_dosatA522_epi.RData
- data/mods_dosatA522_hyp.RData
- data/mods_dosatA522_whole.RData
- data/mods_TSI_chla0611.RData
- data/mods_TSI_chla0612.RData
- data/mods_TSI_chla0804.RData
- data/mods_TSI_chla0826.RData
- data/mods_TSI_chla0831.RData
- data/mods_TSI_chla0852.RData
- data/mods_TSI_chlaA522.RData
- data/mods_TSI_secchi0611.RData
- data/mods_TSI_secchi0612.RData
- data/mods_TSI_secchi0804.RData
- data/mods_TSI_secchi0826.RData
- data/mods_TSI_secchi0831.RData
- data/mods_TSI_secchi0852.RData
- data/mods_TSI_secchiA522.RData
- data/mods_TSI_TP0611.RData
- data/mods_TSI_TP0612.RData
- data/mods_TSI_TP0804.RData
- data/mods_TSI_TP0826.RData
- data/mods_TSI_TP0831.RData
- data/mods_TSI_TP0852.RData
- data/mods_TSI_TPA522.RData
- data/mods_TSI_TN0611.RData
- data/mods_TSI_TN0612.RData
- data/mods_TSI_TN0804.RData
- data/mods_TSI_TN0826.RData
- data/mods_TSI_TN0831.RData
- data/mods_TSI_TN0852.RData
- data/mods_TSI_TNA522.RData
---


```{r, message = F, warning = F, results = 'hide', echo = F}
# globals
options(repos = c(tbeptech = 'https://tbep-tech.r-universe.dev',CRAN = 'https://cloud.r-project.org',USGS = 'https://owi.usgs.gov/R'))

library(knitr)
opts_chunk$set(warning = FALSE, message = FALSE, dev.args = list(bg = 'transparent'), eval = T, echo = F)

library(tidyverse)
library(lubridate)
library(shiny)
# if (!require(wqtrends)) install.packages('wqtrends',repos='https://tbep-tech.r-universe.dev')
# remotes::install_github('tbep-tech/wqtrends')
library(wqtrends)
# devtools::load_all('../wqtrends')
library(ggmap)
library(mgcv)
library(plotly)
library(gridExtra)
library(shinycssloaders)
source('R/funcs.R')

data(datprc)
data(locs)
data(map)

# CLD
#yrs <- seq(1990, 2019)
yrs <- seq(1993, 2021)
# params <- list(
#   'Chlorophyll-a (ug/L)' = 'chl', 
#   'GPP (mg C m-2 d-1)' = 'gpp', 
#   'DO (mg/L)' = 'do', 
#   'DO sat. (%)' = 'dosat'
# )
params <- list(
  'Fecal Coliform (cfu/L)' = 'fc',
    'DO Saturation (%)' = 'dosat',
  'Temperature (â°C)' = 'temp', 
  'Dissolved Oxygen (mg/L)' = 'do', 
  'Conductivity (uS/cm)' = 'cond',
  'pH' = 'ph',
  'Total Phosphorus (mg/L)' = 'tp',
  'Soluble Reactive Phosphorus (mg/L)' = 'srp',
  'Nitrate Nitrogen (mg/L)' = 'no3',
  'Secchi Transparency (m)' = 'secchi',
  'Chlorophyll a (ug/L)' = 'chl',
  'Total Alkalinity (mg/L)' = 'alk',
  'Total Nitrogen (mg/L)' = 'tn',
  'Trophic State Index (Chl a)' = 'TSI_chla',
  'Trophic State Index (Secchi)' = 'TSI_secchi',
  'Trophic State Index (TP)' = 'TSI_TP',
  'Trophic State Index (TN)' = 'TSI_TN',
  'Schmidt Stability (J/m2)' = 'ss'
  
)

# ext <- make_bbox(locs$lon, locs$lat, f = 0.2)
# map <- get_stamenmap(ext,  zoom = 11, maptype = "toner-lite", where = getwd())
pbase <- ggmap(map) +
  theme_bw() +
  theme(
    axis.title.x = element_blank(),
    axis.title.y = element_blank()
  )
```

```{r reactives}
# selected station location and raw time series
mapselplo <- reactive({

  # inputs
  station <- input$station
  parameter <- input$parameter
  typ <- input$typ
  ylb <- ylb()

  req(ylb)
  
  locpt <- locs %>% 
    # filter(Station %in% !!station)
    #CLD
    filter(Station %in% !!substr(station,1,4))

  # map
  p1 <- pbase + 
    geom_point(data = locpt, aes(x = lon, y = lat), colour = 'tomato1', size = 8) +
    geom_text(data = locs, aes(x = lon, y = lat, label = Station))

  # time series raw
  toplo <- datprc %>% 
    filter(station %in% !!station) %>% 
    filter(param %in% !!parameter)
  
  if(typ == 'tot')
     p2 <- ggplot(toplo, aes(x = cont_year, y = value)) + 
        geom_line() 
  if(typ == 'yrs')
     p2 <- ggplot(toplo, aes(x = factor(yr), y = value)) +
            geom_boxplot() 
  if(typ == 'mos')
     p2 <- ggplot(toplo, aes(x = mo, y = value)) + 
        geom_boxplot()
      
  # create plots for whole time series, by year, by month
  p2 <- p2 +
    theme_bw() +
    theme(axis.title.x = element_blank())

  if(typ == 'yrs')
     p2 <- p2 +
    theme(axis.text.x = element_text(angle = 45,hjust=1)) 
  
  # CLD
  # if(parameter %in% c('chl', 'gpp'))
  if(parameter %in% c('chl', 'gpp', 'fc'))
    p2 <- p2 + 
      scale_y_log10(ylb)
  if(parameter %in% c('do', 'dosat'))
    p2 <- p2 + 
      scale_y_continuous(ylb)
  
  out <- grid.arrange(p1, p2, widths = c(0.3, 0.7), ncol = 2)
  
  return(out)
  
})  
  
# models for selected station
mod <- reactive({
  
  # inputs
  parameter <- input$parameter
  station <- input$station
  
  fl <- paste0('mods_', parameter, station)
  load(file = paste0('data/', fl, '.RData'))
 
  out <- get(fl) %>% 
    ungroup() %>%
    pull(modi) %>% 
    deframe()

  return(out)
  
})
  
# y axis label
ylb <- reactive({
  
  # input
  parameter <- input$parameter

  out <- params %in% parameter %>% which %>% params[.] %>% names
  
  return(out)

})

# text labels for parameter selection
txtlb <- reactive({
  
  # input
  parameter <- input$parameter
  
  out <- params %in% parameter %>% which %>% params[.] %>% names
  out <- gsub('^(.*)\\s\\(.*$', '\\1', out)
  out <- case_when(
    out %in% 'Chlorophyll-a' ~ tolower(out), 
    T ~ out
  )
  
  return(out)
  
})

# smoother stats of GAMs for selected station
modtabsmth <- reactive({
  
  # inputs
  mod <- mod()
  
  out <- anlz_smooth(mod)
  
  return(out)
  
})

# summary stats of GAMs, model fit/perf
modtabfit <- reactive({
  
  # input 
  mod <- mod()
  
  out <- anlz_fit(mod)
  
  return(out)
  
})

# doy plot
prddoy <- reactive({
  
  # inputs
  mod <- mod()
  ylb <- ylb()
  
  out <- show_prddoy(mod, ylab = ylb)
  
  return(out)
  
})

# series plot
prdseries <- reactive({
  
  # inputs
  mod <- mod()
  ylb <- ylb()
  
  out <- show_prdseries(mod, ylab = ylb)
  
  return(out)
  
})

# seasonal predictions
prdseason <- reactive({
  
  # inputs
  mod <- mod()
  ylb <- ylb()
  
  out <- show_prdseason(mod, ylab = ylb)

  return(out)
  
})

# 3d plot
prd3d <- reactive({
  
  # inputs
  mod <- mod()
  ylb <- ylb()
  
  out <- show_prd3d(mod, ylab = ylb)
  
  return(out)
  
})

# gam check plot for selected GAM for selected station
gamchk <- reactive({
  
  # inputs
  mod <- mod()
  
  out <- mod %>% 
    gam.check()
  
  return(out)
  
})

# percent change
perchg <- reactive({
  
  # inputs
  yrbs <- as.numeric(input$yrbs)
  yrts <- as.numeric(input$yrts)
  mod <- mod()
  ylb <- ylb()
  
  req(yrbs)
  req(yrts)
  
  out <- show_perchg(mod, baseyr = yrbs, testyr = yrts, ylab = ylb)
  
  return(out)
  
})

# seasonal averages
metseason <- reactive({
  
  # inputs
  dytr <- input$dytr
  yrtr <- input$yrtr
  metsel <- input$metsel
  mod <- mod()
  ylb <- ylb()
  
  req(yrtr)
  
  # add explicit quotes for parse
  ylb <- paste0('\"', ylb, '\"')

  out <- try({
    
    toprs <- paste0('show_metseason(mod, metfun = ', metsel, ', doystr = ', dytr[1], ', doyend = ', dytr[2], ', yrstr = , ', 
                    yrtr[1], ', yrend = ', yrtr[2], ', ylab = ', ylb, ', nsim = 1e3, na.rm = TRUE)')
    
    eval(parse(text = toprs))
    
    })
  
  validate(
    need(inherits(out, 'ggplot'), 'Pick different year range')
  )
  
  return(out)
  
})

# seasonal trends
trndseason <- reactive({
  
  # inputs
  dytr <- input$dytr
  yrtr <- input$yrtr
  mod <- mod()
  metsel <- input$metsel
  wntr <- input$wntr
  wnty <- input$wnty
  
  req(yrtr)
  
  # add 
  out <- paste0('show_trndseason(mod, ', metsel, ', doystr = ', dytr[1], ', doyend = ', dytr[2], ', win = ', wntr, 
    ', justify = \"', wnty, '\", ylab = \"Change per year\", nsim = 1e3, na.rm = TRUE)')
 
  out <- eval(parse(text = out))
  
  return(out)
  
})

```

This application uses data from long-term water quality monitoring of large lakes of the Seattle metropolitan area (lakes Union, Washington, and Sammamish) ([King County 2020](https://green2.kingcounty.gov/ScienceLibrary/Document.aspx?ArticleID=624){target="_blank"}) and the [wqtrends](https://tbep-tech.github.io/wqtrends/){target="_blank"} package in R.  Details on the GAM-mixmeta approach are presented in [Beck et al (2022)](https://doi.org/10.1016/j.scitotenv.2021.149927){target="_blank"} and the original San Francisco Bay Trends dashboard from which this application was developed can be found [here](https://nutrient-data.sfei.org/apps/SFbaytrends/){target="_blank"}. The github repository for the San Francisco Bay R source code that provided the basis for this application can be found [here](https://github.com/fawda123/SFBaytrends/){target="_blank"}.

## Exploratory plots

The following plots show the raw data for long-term lake monitoring stations and parameters, 1993 - 2021.  Select the station, parameter, and plot type (total time series, by year, or by month).  The year and month plots are aggregated box plots of all observations at a station for each selected time period.

```{r, echo = F}
# selection widgets
column(4, selectInput("station", "Choose station:", sort(unique(datprc$station))))
column(4, selectInput("parameter", "Choose parameter:", choices = params))
##CLD 
# tst <- keep(params, ~ any(pms %in% .x))
# column(4, selectInput("parameter", "Choose parameter:", choices = tst))
column(4, selectInput("typ", "Choose plot type:", c('tot', 'yrs', 'mos')))

## CLD
observe({
  pms <- unique(datprc$param[datprc$station %in% input$station])  
  newparameter <- purrr::keep(params, ~ any(pms %in% .x))

  updateSelectInput(session, "parameter",
                  choices = newparameter)

})


```

```{r}
# make the plot
output$mapselplo <- renderPlot(mapselplo(), height = 300, width = 900)
plotOutput('mapselplo') %>% withSpinner()
```

## GAM results {.tabset}

Generalized additive models (GAMs) were developed to describe trends at each of the monitoring stations.  The station and parameter can be chosen from the drop down menus above.  Each model has the following structure:

* `S`: `r renderText(input$parameter)` ~  s(cont_year, k = *large*)

The `cont_year` variable is measured as a continuous numeric value for the time series vector (e.g., January 1st, 2000 is 2000.0, July 1st, 2000 is 2000.5, etc.). The function `s()` models `year` as a smoothed, non-linear variable.  The `k` value is a theoretical upper limit on the number of knots used to fit the smoother for `cont_year`.  The optimal number of knots are determined automatically using generalized cross-validation as implemented with the `mgcv()` function.  Using a sufficiently large number for `k` ensures the model is not underfit to the data. The upper limit of `k` was chosen as 12 times the number of years of data at each station (~360).

__Selected station:  `r renderText({input$station})`, Selected parameter: `r renderText(txtlb())`__

### Predictions

The four plots show the predicted trends from the model.  The first plot shows estimated `r renderText(txtlb())` by day of year with separate lines for each year.  The second plot show predictions across the entire time series. The third plot show the predicted trends with separate lines for each month across the time series. The final plot shows the results as a three-dimensional fit of the estimated trends across year and day of year with the z-axis showing estimated `r renderText(txtlb())`.  

```{r}
output$prddoy <- renderPlot(prddoy())
plotOutput('prddoy', height = "400px", width = "900px") %>% withSpinner()
```

```{r}
output$prdseries <- renderPlot(prdseries())
plotOutput('prdseries', height = "330px") %>% withSpinner()
```

```{r}
output$prdseason <- renderPlot(prdseason())
plotOutput('prdseason', height = "400px") %>% withSpinner()
```

```{r}
output$prd3d <- renderPlotly(prd3d())
plotlyOutput('prd3d') %>% withSpinner()
```

### Fit

The first table shows the estimates for the continuous year smoother as the estimated degrees of freedom (edf), the reference degrees of freedom (Ref.df), the test statistic (F), and significance of the component (p-value). The significance is in part based on the difference between edf and Ref.df. The second table shows the overall model summary as Akaike Information Criterion (AIC), the generalized cross-validation score (GCV), and the R2 values. In general, lower values for AIC and GCV and higher values for R2 indicate improved model fit. The plots are results from the `gam.check()` function that provides a standard assessment of GAM fit.

```{r}
output$modtabsmth <- renderTable(modtabsmth())
output$modtabfit <- renderTable(modtabfit())
tableOutput('modtabsmth')
tableOutput('modtabfit')
```

```{r}
output$gamchk <- renderPlot(gamchk(), height = 550)
plotOutput('gamchk', height="550px") %>% withSpinner()
```

### Trends 

This tab shows the predicted trends across the time series (and options for trend assessment) and the predicted trends as seasonal aggregations with 95% confidence intervals within each year.

```{r}

column(4,
       renderUI({

         req(mod())
         yrs <- mod()$model %>% pull(cont_year) %>% date_decimal(.) %>% year %>% unique %>% sort()
         selectInput('yrbs', 'Select base years:', choices = yrs, selected = yrs[1], multiple = T)
         
         })
)
column(4,
       renderUI({
         
         req(mod())
         yrs <- mod()$model %>% pull(cont_year) %>% date_decimal(.) %>% year %>% unique %>% sort()
         selectInput('yrts', 'Select test years:', choices = yrs, selected = rev(yrs)[1], multiple = T)
         
         })
)
```

```{r}
output$perchg <- renderPlot(perchg(), height = 320)
plotOutput('perchg') %>% withSpinner()
```

```{r}
column(12,
  column(6,
    ##CLD sliderInput('dytr', 'Select day-of-year range to define seasons:', min = 1, max = 365, value = c(213, 304))
    sliderInput('dytr', 'Select day-of-year range to define seasons:', min = 1, max = 365, value = c(152, 273))
  ),
  column(6,
    renderUI({
      
      req(mod())
      yrs <- mod()$model %>% pull(cont_year) %>% date_decimal(.) %>% year %>% unique
      # sliderInput('yrtr', 'Select years for across-year trend estimate:', min = min(yrs), max = max(yrs), value = c(2015, 2019), sep = "", step = 1)
      # CLD
      sliderInput('yrtr', 'Select years for across-year trend estimate:', min = min(yrs), 2025, value = c(2015, 2021), sep = "", step = 1)
      
    })
  )
)
column(12, 
  column(6, 
    selectInput('metsel', 'Select summary metric:', choices = c('mean', 'min', 'max', 'var'), selected = 'mean')       
  )
)
```

```{r}
output$metseason <- renderPlot(metseason(), height = 320)
plotOutput('metseason') %>% withSpinner()
```

<br></br>
<br></br>
<br></br>

```{r}
column(12, 
  column(6, 
    sliderInput('wntr', 'Select window size:', min = 2, max = 20, value = 5)       
  ), 
  column(6, 
    selectInput('wnty', 'Select window type:', choices = c('center', 'left', 'right'))     
  )
)
```

```{r}
output$trndseason <- renderPlot(trndseason(), height = 370)
plotOutput('trndseason') %>% withSpinner()
```

